---
title: "GIS for Transport Applications"
author: "Robin Lovelace"
date: "April 14, 2015"
output:
  ioslides_presentation:
    transition: slower
---

```{r, echo=FALSE, include=FALSE}
pkgs <- c("grid", "png")
lapply(pkgs, library, character.only = T)
# bibliography: "~/Documents/R.bib"
```

## Welcome and housekeeping

09:30 – 10:00   Registration and refreshments

10:00 – 11:00	 Introduction to R and QGIS for transport applications

11:00 – 12:00	 Route analysis using R and Routino 

12:00 – 12:30	 Presentation of maps 

**12:30 – 13:30 	 Lunch**

13:30 – 14:30	Collecting, processing and analysing large GPS and Smartphone datasets

14:30 – 16:30	 Hackathon 

16:30 – 17:00	 Best map prize announcement.

# QGIS for transport applications

## The uptake of QGIS

```{r, echo=FALSE}
grid.raster(readPNG("/home/robin/Pictures/Selection_178.png"))
```


## Flexible and user friendly

```{r, echo=FALSE}
grid.raster(readPNG("~/repos/bikeR/figures/wy-overview.png"))
```

## Compared with R...

```{r, echo=FALSE}
grid.raster(readPNG("~/repos/bikeR/figures/seriousWY.png"))
```

# R for transport applications

## Why R?

See the free, open source, online tutorial [github.com/robinlovelace/Creating-maps-in-R](https://github.com/Robinlovelace/Creating-maps-in-R).

```{r, echo=FALSE}
f <- "/home/robin/repos/Creating-maps-in-R/figure/facet_london.png"
grid.raster(readPNG(f))
```

## The packages used

Can be installed and loaded in 6 lines of code:

```{r, eval=FALSE}
pkgs <- c("devtools", "rgdal", "raster", "ggmap") # official packages
install.packages(pkgs) 
library(devtools) # enables installation of leaflet
gh_pkgs <- c("rstudio/leaflet", "robinlovelace/stplanr") 
install_github(gh_pkgs) # install packages on github
lapply(c(pkgs, "leaflet", "stplanr"), library, character.only = T) # load all
```

## RStudio

RStudio Desktop is highly recommended.

```{r, echo=FALSE}
img <- readPNG("/home/robin/repos/robinlovelace.github.io/img/rstudio-shiny-button.png")
grid.raster(img)
```

```{r, echo=FALSE, include=FALSE}
pkgs <- c("devtools", "shiny", "rgdal", "rgeos", "ggmap") # official packages
gh_pkgs <- c("rstudio/leaflet", "robinlovelace/stplanr") 
lapply(c(pkgs, "leaflet", "stplanr"), library, character.only = T) # load all
```

## Leaflet

An R interface to the Leaflet JavaScript library, compatible with Shiny.

```{r, eval=FALSE}
cent <- geocode("Leeds")
leaflet() %>% 
  addTiles() %>%
  addCircleMarkers(data = cent)
```

# III: National Propensity to Cycle Tool (NPCT)

## Early user testing

```{r, echo=FALSE}
grid.raster(readPNG("~/Pictures/Selection_115.png"))
```

## Genius additions: zoom dependent selection; freeze Scope

```{r, echo=FALSE}
grid.raster(readPNG("~/Pictures/Selection_156.png"))
```

## Running the app for different cities

```{r, eval=FALSE}
runApp("~/repos/pct-shiny/norwich/", launch.browser = T)
runApp("~/repos/pct-shiny/cambridge/", launch.browser = T)
runApp("~/repos/pct-shiny/manchester/", launch.browser = T)
```

# IV: Get your hands dirty

## Working through **Introduction to R and QGIS**

- Don't worry if you don't finish
- Prioritise either R or QGIS
- Interact with us and others
- Any questions

