---
title: "Introduction to R and QGIS for transport applications"
author: "Robin Lovelace"
output:
  pdf_document:
    fig_caption: yes
---

```{r, echo=FALSE}
library(grid)
library(png)
```


# Introduction

This document provides a taster of QGIS and R, two of the most popular
and powerful open source GIS productions, for transport applications.
Both programs are vast, in terms of features and their code-base,
and take many months of dedicated practice to master. Therefore
the aim is to showcase capabilities and demonstrate key functionality.
This will hopefully form the foundation for further study; the
references provide some pointers in that direction.

Conventional GIS courses are rather 'spoon feed'
(see [ESRI's free online courses](http://www.esri.com/training/main the students), for example), providing a long list of screenshots to follow
in a set order. This course takes a deliberately different approach.
The challenges are deliberately more open-ended and you are encouraged
to search on-line, explore the software's help and, most importantly,
ask each other (and the amazing FOSS4G community) for help.

We will begin by using QGIS because it is easier to get used to for most
people, with it's graphical user interface. R, by contrast, has a steep
learning curve if you are not acquainted with using a command-line interface.
First, let's download the data we'll be using for this tutorial.

# The data

The datasets used for this project can be downloaded in a single file,
from the [GIS4TA-data repository](https://github.com/Robinlovelace/GIS4TA-data). The [ZIP file](https://github.com/Robinlovelace/GIS4TA-data/archive/master.zip) is automatically 
created by GitHub and is available for download from here:
https://github.com/Robinlovelace/GIS4TA-data/archive/master.zip

All the data we'll be using in this tutorial is in the `leeds`
directory. In this folder you should see the following files:

- `cyclepaths.geojson`, representing dedicated cycle paths in Leeds
- `msoas-slope.*`, these files together constitute a 'Shapefile' 
representing the 107 'MSOA' administrative zones of Leeds Local Authority.
- `travel-net-mini.sqlite`, a dataset representing the transport network of
Leeds.

These will form the basis of our analysis in QGIS and R.


# QGIS

QGIS is the most widely used open source GIS program and an amazing
example of community-led software development. It is a general-purpose
geographic information system that can solve a wide range of geographical
problems. With its interface to Python, it has a strong user community
developing powerful, attractive and user-friendly add-ons.

The best way to learn more about QGIS is to use it.
Download the latest version from [qgis.org](http://www.qgis.org). 

For this chapter we will primarily be using the basic QGIS
features to load, explore and query geographic information related
to transport systems. However, there are many
plugins suitable for transport applications.
These can be installed from
**Plugins > Manage and Install Plugins**. In this chapter we'll be using
the following two plugins:

- The **OpenLayers Plugin** is not transport-specific but provides an
interactive, 'slippy' base map on which transport data can be displayed.
In addition, a number of transport-orientated maps are provided, including
*OpenCycleMap*.
- **Road Graph** calculates the shortest path between origins and
destinations.

In addition to these (needed for the tutorial) the following two plugins
are useful for more advanced applications:

- **FlowMapper** provides functionality for visualising travel flows
between origins and destinations (a flow matrix).
- **PGRouting**, an interface to the powerful **PGRouting** extension to
the PostGIS spatial database system for route planning.

We will not be using these for the present chapter. For now we focus on
getting to grips with basic tasks such as loading and saving data.

## Loading and saving data

QGIS can load and save data in a wide variety of formats.
To load the administrative zones, open the 'msoas-slope' Shapefile.
Shapefiles are a very common proprietary GIS format gradually
being replaced by alternatives such as
[GeoJSON](http://geojson.org/) (and the related
[TopoJSON](https://github.com/mbostock/topojson)).

Click on the 'add vector layer' button on the left and navigate
to the data folder. Double clicking on the `msoas-slope.shp`
should load the file into QGIS.

**Challenge: add a transport-related basemap using the OpenLayers plugin loaded in the first section.**

The network data was downloaded using the **OpenStreetMap**
Plugin. The following code was entered into the filtering panel to 
remove the superfluous tracks:

```
"highway"  != 'unclassified'  AND "highway" != 'unsurfaced' 
AND "highway"  != 'bridleway'  AND "highway" != 'living_street'
AND "highway"  != 'tertiary'  AND "highway" != 'track'
```

We've already saved the data for you, so you don't need to
go through this process. The point is that
with open source it is easy to document a dataset's origin.

To load the transport network, use the 'Add Spatialite Layer' option.

**Challenge: load and look at this transport layer**

## Finding the shortest path in QGIS

Activate the Road Graph plugin and you should see a new panel appear
in the left-hand side of the screen, below 'Layers'. Experiment
with this. See if you can find the fastest route across Leeds.
the end result should look a little like that displayed in 
Fig. 1.

```{r, fig.cap="A shortest part across Leeds, generated by QGIS", echo=FALSE}
grid.raster(readPNG("figures/leeds-shortest.png"))
```

**Challenge: find the shortest path from Leeds station to the University in QGIS.**

## Subset and save data using QGIS

There are a number of ways to subset data from QGIS. You can select
with a box or other polygon, use a spatial intersect with another
object or use the the 'filter tool'. One of the most versatile tools
available to do this (and a great way to view the attribute data of
spatial objects) is the the Attribute Table.

Following the steps below will result in the creation of a new object,
a *subset* of network data, that removes everything other than 
the type of route you desire.

1. Right click on the travel_net layer in the layers panel and select  **Filter**.
2. Select first the variable of interest and then the value
(clicking on 'sample' will show a selection of these). See Fig. 2.
The colour of the selected features should change to yellow.
3. To save the subset of the data, again right click on the layer name
and select **Save**. Critically, click **Save only selected features**.
There are many formats to choose from. We used the GeoJSON output type
for the cycle paths. These will be used in the subsequent section.

```{r, fig.cap="The query builder interface in QGIS", fig.height=3}
grid.raster(readPNG("figures/query-builder.png"))
```

## Exporting a map of an area of interest

The final QGIS challenge is open-ended. Identify a place you are
particularly interested with reference to a (real or imaginary)
transport problem. The challenge is to create a map for export
that illustrates the problem (hint: use the 'print composer').
Bonus points if your map can provide insight into the solution.

# R for transport applications

R is a powerful and extremely flexible program, with over 
7,000 add-on packages for solving all manner of numerical
(and some non-numerical) problems.

We will only scratch the surface of what is possible with R:
it's better to gain a good understanding of the basics and
progressing gradually than going instantly to advanced applications.

So in this section we will simply load and subset the data.
We will use the **raster** and **RSQLite** packages for loading the
Shapefile and SQLite datasets respectively.

From within RStudio, enter the following commands in the console:

```{r, eval=FALSE}
pkgs <- c("raster", "rgdal")
install.packages(pkgs)
lapply(pkgs, library, character.only = T)
```

```{r, include=FALSE}
pkgs <- c("RSQLite", "raster", "devtools", "rgdal")
lapply(pkgs, library, character.only = T)
```


We'll start by loading and viewing the Shapefile:

```{r, fig.show='hide'}
leeds_msoas <- shapefile("GIS4TA-data/leeds/msoas-slope.shp")
plot(leeds_msoas)
```

If an image of Leeds Local Authority appeared, congratulations: you've
just loaded and plotted some spatial data in R!

```{r, fig.show='hide'}
leeds_net <- readOGR(dsn = "GIS4TA-data/leeds/cyclepaths.geojson", layer = "OGRGeoJSON")
plot(leeds_net)
```

The final challenge is to use a basic filter to find canal paths.
To see what variables are stored in the cycle paths,
type `names(leeds_net)`.
Further, explore the `highway` types and modify the code below to
plot different types of path:

```{r, fig.height=2.5, fig.cap="Aire Valley Towpath plotted with R"}
summary(leeds_net$name)[1:3] # some of the line names
sel <- grep(pattern = "Aire Valley Towpath", x = leeds_net$name)
towpaths <- leeds_net[sel, ] # create a new object of towpaths
plot(towpaths) # plot the result
```

There are many more things we can do with R for transport applications.
Because it is an extremely flexible programming language, with strong
interfaces to all major low-level languages such as C++, you really are
limited only by your imagination with R if you can program. Fortunately,
even if you cannot, R's extensive range of packages provides functions
useful for transport planners. Moreover, the supportive open-source community
surrounding R has provided a number of resources that enable
further learning. Some of these are described in the further resources and
references. In addition, online sources such as
[r-bloggers](http://www.r-bloggers.com/),
[stackoverflow](http://stackoverflow.com/)
and [gis.stackexchange](http://gis.stackexchange.com/)
are recommended.

# References

Graser, A. (2014). Learning Qgis. Packt Publishing Ltd.

Lovelace, R. (2014). Harnessing open street map data with R and QGIS. EloGeo. Retrieved from http://eprints.whiterose.ac.uk/77643/

Lovelace, R., & Cheshire, J. (2014). Introduction to visualising spatial data in R. National Centre for Research Methods Working Papers, 14(03). Retrieved from https://github.com/Robinlovelace/Creating-maps-in-R

Urrutia Fernández, M. (2014). Quantum GIS (QGIS) Geographic Information System Tutorial. http://upcommons.upc.edu/pfc/bitstream/2099.1/23860/1/TFM_MLTM%20M.Angels%20Urrutia_english.pdf



```{r, eval=FALSE, echo=FALSE}
install_github("pschmied/RSQLite.spatialite")
drv <- dbDriver("SQLite")
con <- dbConnect(drv = drv, dbname = "GIS4TA-data//leeds/travel-net.sqlite")
init_spatialite(con)
tnet <- dbGetQuery(con, "SELECT *")
```

