---
title: "Open Street Map for Transport Applications"
author: 
  - name: "Mark Padgham"
    url: https://github.com/mpadge
date: 2017-11-17
output: 
    html_document:
        theme: flatly
---

```{r, echo=FALSE}
knitr::opts_knit$set(root.dir = "..")
```

# development versions

Much of this code uses the latest 'development' versions of software packages.
These are not what you get from a standard `install.packages()` in R, but can be
easily installed directly from github source code with the `devtools` package.
to ensure everything works as intended, it will likely help to run the following
lines
```{r}
install.packages ("devtools")
library (devtools)
install_github("tidyverse/ggplot2")
install_github("ropensci/osmdata")
install_github("ATFutures/dodgr")
```

# bristol data

The street network for Bristol can be obtained with these lines:
```{r}
library (osmdata)
dat <- opq ("bristol") %>%
    add_osm_feature (key = "highway") %>%
    osmdata_sf ()
```
It's probably wise to save these data for future re-use:
```{r}
save (dat, file = "osm-bristol.rda")
```
They can then be re-loaded with
```{r}
load ("osm-bristol.rda")
```

## Viewing data#1

These data can not be directly viewed using any current spatial visualisation
packages. The following code simply fails:
```{r}
library (mapview)
mapview (dat$osm_lines)
```

This code works, but generates rather visually unappealing results, and is
notably slow. (Execution of these lines is not recommended.)
```{r}
library (ggplot2)
library (sf)
dat <- dat$osm_lines
g <- ggplot () + geom_sf(data = dat)
ggsave (g, file = "bristol-ggplot.eps")
```

# dodgr

`dodgr` converts `sf`-format line data into a simple tabular network.
```{r}
library (dodgr)
net <- weight_streetnet (dat$osm_lines, wt_profile = "bicycle")
```
The data can be examined with:
```{r}
head (net)
dim (net)
```

## dodgr distances

An example of calculating distances between two points:
```{r}
from <- data.frame (x = -2.7, y = 51.4)
to <- data.frame (x = -2.6, y = 51.5)
dodgr_dists (net, from = fr, to = to)
```

## dodgr flows

To calculate flows, we need to specify the vertices of the network in terms of
their `id` values (as defined by OSM).

```{r}
verts <- dodgr_vertices (net)
head (verts)
```
The following lines randomly select `npts = 100` of these:
```{r}
npts <- 100
pts <- sample (verts$id, size = npts)
```
A 'flow' layer is then simply a matrix with numbers of rows equal to the length
of `from` points and numbers of columns equal to the length of `to` points. In
this case, it's a square matrix with both of these equal to `npts`.
```{r}
flows <- matrix (1, nrow = npts, ncol = npts)
```
The actual flows can then be calculated with:
```{r}
f <- dodgr_flows (net, from = pts, to = pts, flows = flows) %>%
        merge_directed_flows ()
```
The `merge_directed_flows` removes all edges with zero flow, and so yields a
more manageable table for subsequent manipulation and visualisation.

## viewing data#2

The following code generates a visual representation of the flows.
```{r}
require (osmplotr)
require (ggplot2)
bb <- getbb ("bristol")
f$flow <- f$flow / max (f$flow)
cols <- colorRampPalette (c ("dodgerblue", "firebrick")) (30)
map <- osm_basemap (bb, bg = "#FFEEBB") %>%
    add_colourbar (zlims = range (f$flow), col = cols)
map <- map +
    geom_segment (aes (x = from_lon, y = from_lat, xend = to_lon, yend = to_lat,
                       colour = flow, size = flow), size = 3 * f$flow,
                  data = f) +
    scale_colour_gradient (low = "dodgerblue", high = "firebrick", guide = "none")
```
Typing `map` will open a viewing window, but rendering may be very slow for
large graphs, so it may be quicker to plot to a file and view that separately:
```{r}
print_osm_map (map, file = "bristol-flows.png")
```
(The function `print_osm_map()` primarily ensures that the map is printed in
proportion to the bounding box.)

Finally, the following function can be used to generate flow values between
differing numbers of random routing points. This illustrates the kinds of ways
that `dodgr` can be used to calculate flows.
```{r}
make_bristol_map <- function (net, npts = 100)
{
    pts <- sample (verts$id, size = npts)
    flows <- matrix (1, nrow = npts, ncol = npts)
    f <- dodgr_flows (net, from = pts, to = pts, flows = flows) %>%
        merge_directed_flows ()
    f$flow <- f$flow / max (f$flow)
    cols <- colorRampPalette (c ("dodgerblue", "firebrick")) (30)
    map <- osm_basemap (bb, bg = "#FFEEBB") %>%
        add_colourbar (zlims = range (f$flow), col = cols)
    map <- map +
        geom_segment (aes (x = from_lon, y = from_lat, xend = to_lon, yend = to_lat,
                           colour = flow, size = flow), size = 3 * f$flow,
                      data = f) +
            scale_colour_gradient (low = "dodgerblue",
                                   high = "firebrick", guide = "none")
    fname <- paste0 ("bristol-flows", npts, ".png")
    print_osm_map (map, file = fname)
}
```
A map can then be produced with,
```{r}
make_bristol_map (net, 100)
```
